---
title: "Code Through Tutorial on 'lubridate()'"
author: "Alev Yildiz"
date: "10/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Introduction to 'lubridate()' package**

- The purpose of this code through tutorial is to show how to work with dates and times in R. At first glance, dates and times may seem simple; however, the more you learn about them, the more complicated they get.
- Lubridate package makes it easier to identify and parse date-time data, extract and modify components of a date-time, and handle time zone. 

# **Creating Date/Times**

- Let's look at the three ways to create a date/time:
  - Strings
  - Individual Components
  - Date or Time Components
  
#### Before we go more in depth, let's import necessary packages. 
```{r echo=FALSE}
library(tidyverse)
library(lubridate)
library(nycflights13)
library(ggplot2)
library(scales)
```

- Three types of date/time data which refers to an instant time: A _date_ <date>; A _time_ within a day <time>; and a _date-time_ is a date plus time. 
- 'lubridate()' allows us to  find the days of the week, month and more within an inputted date.

```{r}
weekdays(today()) #Thursday
months(today())   #October
quarters(today()) #Q4
year(today())     #2021
```

<br>
<br>

# **Strings**

- Date/time data often comes as strings. Using the helpers provided by lubridate automatically creates the format once you specify the order of the component. To use them, create the order in which year, month and day appear in your dates, then arrange "y", "m", and "d" in the same order. For example:

```{r}
mdy( "October 16th, 2020")
ymd("2020-08-27")
dmy("15-Dec-2020")
```

<br>
<br>

- We can also use these functions with unquoted numbers, which may be needed when filtering date/time data. 

```{r}
ymd(19950102)
```

<br>
<br>

- The creation of a date-time from a date by supplying a time zone:
```{r}
ymd(20120101, tz= "UTC")
```

<br>
<br>

# **From Individual Components (using 'flights' data)**

- Instead of creating a date/time data with a single string, we will have the individual components of the date-time spread across multiple columns. Let's explore the flights data:

```{r}
#The date-time spread across multiple columns
#flights %>% select(year, month, day, hour, minute)
```

<br>
<br>

- Use make_date() for dates, or make_datetime() for date-times to create a date/time from the input above. 

```{r echo=FALSE}
flights %>% select(year, month, day, hour, minute) %>%
mutate(departure= make_datetime( year, month, day, hour, minute))
```

<br>
<br>

# **Date-Time Components**

**We can pull out individual parts of the date with the following functions: year(), month(), mday() (day of the month), yday() (day of the year), wday() (day of the week), minute(), and second():**

```{r}
date.time <- ymd_hms("2018-08-08 01:20:44")
year(date.time) #2018
month(date.time) #08
mday(date.time) #08

yday(date.time) #220
wday(date.time) #4

```

<br>
<br>

**We can abbreviate the name of the month or day of the week** 
```{r}
month(date.time, label = TRUE)
#Jul
wday(date.time, label=TRUE, abbr = FALSE)
#Wednesday
```

<br>
<br>

# **We can use each accessor function to set the components of a date/time:**

```{r}
(datetime <- ymd_hms("2012-02-05 10:05:45"))
# "2012-02-05 10:05:45 UTC"

year(datetime) <- 2015
datetime
#"2015-02-05 10:05:45 UTC"
month(datetime) <- 12
datetime
#"2015-02-05 10:05:45 UTC"
hour(datetime) <- hour(datetime) + 2
datetime
#"2015-02-05 12:05:45 UTC"
```

<br>
<br>

**lubridate package performs artihmetic calculation on date-times.** 

-Let's assume that I was born on January 02, 1968. How old am I?
```{r}
myage <- today() - ymd(19680102)
myage
#The output I get for my age is: "Time difference of 19637 days". Apparently I am 19637 days. 
#Lubridate provides an alternative that uses seconds-the duration:
as.duration(myage)
#"1696636800s (~53.76 years)" That sounds much better!
```

<br>
<br>

**'lubridate()' package has convenient functions for manipulating dates. This can be extremely helpful when we are creating a line plots which displays a metric over the course of a year for many years.**

```{r echo=FALSE}
library(ggplot2)

## let's create year and month variables
economics$year <- year(economics$date)
#the label argument to month indicates that the result should #be the names of the month instead of the number
economics$month <- month(economics$date, label = TRUE)

# subset the data to include only years starting with 2000
econ2000 <- economics[which(economics$year >= 2000), ]
```

<br>
<br>

-Lets create the line plots:

```{r}
#building the foundation of the plot
g<- ggplot(econ2000, aes(x=month, y=pop))
#add lines color coded and grouped by year
g <- g + geom_line(aes(color=factor(year), group=year))
g <- g + scale_color_discrete(name="Year")
g <- g + scale_y_continuous(labels = comma)
g <- g + labs(title="Population Growth", x="Month", y="Population")
g
                   
```
<br>
<br> 

**This package comes in handy when we are dealing with identifying or parsing date-time data and extracting and modifying components of a date-time components.**

# Credits

1. Lander, P. J. (2017). R for Everyone: Advanced Analytics and Graphics. Pearson Education.
2. Wickham H., & Grolemun G. (2017). R for Data Science. O'Reilly Media. 